Model-based pretreatment for regression with correlated variables.

### 2014-04-07 www/article13.tex
tradeoff updated
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/correg@503 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

	The marginal regression model simply is
	\begin{equation}
		\boldsymbol{Y}_{|\boldsymbol{X_1}}= \boldsymbol{X_1}\boldsymbol{\beta_1^*}+ \boldsymbol{\varepsilon_Y^*}
	\end{equation}			
		So we have the \textsc{OLS} unbiased estimator of $\alpha$: 
		\begin{equation}
			\boldsymbol{\hat\beta_1^*} = (\boldsymbol{X_1}' \boldsymbol{X_1})^{-1}\boldsymbol{X_1}'\boldsymbol{Y}  \textrm{ and }\boldsymbol{\hat\beta_2^*} = \boldsymbol{0}
		\end{equation}
		We see in (\ref{Trueexpl}) that it gives an unbiased estimation of $\boldsymbol{Y}$ and $\boldsymbol{\beta^*}$
		but in terms of $\boldsymbol{\beta}$ this estimator is biased:
		\begin{equation}
			\operatorname{E}[\boldsymbol{\hat\beta_1^*}|\boldsymbol{X_1}]=\boldsymbol{\beta_1}+\sum_{j \in I_2}\beta_{j}\boldsymbol{\alpha_j} \textrm{ and }\operatorname{E}[\boldsymbol{\hat\beta_2^*}|\boldsymbol{X_1}]=\boldsymbol{0}
		\end{equation}
		with variance:
		\begin{equation}
			\operatorname{Var}[\boldsymbol{\hat\beta_1^*}|\boldsymbol{X_1}]= (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 )(\boldsymbol{X_1}' \boldsymbol{X_1})^{-1}  \textrm{ and }\operatorname{Var}[\boldsymbol{\hat\beta_2^*}|\boldsymbol{X_1}]= \boldsymbol{0} 
		\end{equation}
		We see that the variance is reduced compared to OLS described in equation (\ref{eqOLS})(no correlations and smaller matrix give better conditioning ) for small values of $\sigma_j$ $i.e.$ strong correlations. So we play on the bias-variance tradeoff, reducing the variance by adding a bias.		
			%There is no theoretical guarantee that our model is better. It's just a compromise between numerical issues caused by correlations for estimation and selection versus increased variability due to structural hypothesis. Therefore we made some simulations to compare both methods (see the end of this paper).
			 \\
	\subsection{Illustration of the tradeoff conveyed by the pretreatment}		  
	 The Mean Squared Error (\textsc{MSE}) on $\boldsymbol{\hat{\beta}}$ is:
	\begin{eqnarray}
		\textsc{MSE}(\boldsymbol{\hat{\beta}}|\boldsymbol{\beta},\boldsymbol{X})&=&\operatorname{Bias}^2+\operatorname{Tr}(\operatorname{Var}(\boldsymbol{\hat{\beta}})) \\
			&=&0+\sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}) \textrm{ for OLS and } 
	\end{eqnarray}	 
	To better illustrate the bias-variance tradeoff, we look at a simple example with $p=3$ variables. $X_1$ is composed by two independent scaled Gaussian $\mathcal{N}(0,1)$, $X_3=x_1+x_2+\varepsilon_3$ where $\varepsilon_3\sim{\mathcal{N}(0,\sigma_3^2)}$. We also have $\beta=(1,1,1)$ and $\sigma_Y=10$. Then we observe the theoretical Mean Squared Error (MSE) of the estimator of both OLS and \textsc{CorReg} model for several values of $\sigma_3$ (strength of the sub-regression) and $n$. Figures \ref{MQE1} to \ref{MQE3} show the theoretical MSE evolution with the strength of the sub-regression:
	\begin{equation}
		1-\mathcal{R}^2=\frac{\operatorname{Var}(\varepsilon_3)}{\operatorname{Var}(x_3)}=\frac{\sigma_3^2}{\sigma_3^2+2}
	\end{equation}

\begin{figure}[h!]
	\begin{minipage}[l]{.32\linewidth}
			\includegraphics[width=170px]{figures/MQEn15sigmaY10.png} 
			\caption{For $n=15$. Dotted: \textsc{Correg}, plain: OLS}\label{MQE1}
	\end{minipage} \hfill
	\begin{minipage}[c]{.32\linewidth}
			\includegraphics[ width=170px]{figures/MQEn100sigmaY10.png} 
			\caption{For $n=100$. Dotted: \textsc{Correg}, plain: OLS}
	\end{minipage} \hfill
   \begin{minipage}[r]{.32\linewidth}
			\includegraphics[width=170px]{figures/MQEn1000sigmaY10.png} 
			\caption{For $n=1000$. Dotted: \textsc{Correg}, plain: OLS.} \label{MQE3}
   \end{minipage} 
\end{figure} 
It is clear in Figures \ref{MQE1} to \ref{MQE3} that the marginal model is more robust than \textsc{OLS} on $X$. And when sub-regression get weaker ($1-\mathcal{R}^2$ tends to 1) it does not explode until extreme values (sub-regression nearly fully explained by the noise). We also see that the error implied by strong correlations shrinks with the rise of $n$.  $MSE_{OLS}=0+\operatorname{Tr}(\operatorname{Var}(\hat{\beta}))=\operatorname{Tr}(\operatorname{Var}(\hat{\beta}_{1}))+\operatorname{Tr}(\operatorname{Var}(\hat{\beta}_{2}))$ where both terms are multiplied by $\sigma_Y^2$ and so is $\operatorname{Tr}(\operatorname{Var}(\hat{\alpha}_{1}))$. Thus, as $\operatorname{Var}(\hat{\alpha}_{2})=0$ when $\sigma_Y^2$ rises it increases the advantage of \textsc{CorReg} versus \textsc{OLS}. It illustrates the importance of dimension reduction when the model has a strong noise (very usual case on real datasets where true model is not even exactly linear).

>>>>

	The marginal regression model simply is
	\begin{equation}
		\boldsymbol{Y}_{|\boldsymbol{X_1}}= \boldsymbol{X_1}\boldsymbol{\beta_1^*}+ \boldsymbol{\varepsilon_Y^*}
	\end{equation}			
		So we have the \textsc{OLS} unbiased estimator of $\alpha$: 
		\begin{equation}
			\boldsymbol{\hat\beta_1^*} = (\boldsymbol{X_1}' \boldsymbol{X_1})^{-1}\boldsymbol{X_1}'\boldsymbol{Y}  \textrm{ and }\boldsymbol{\hat\beta_2^*} = \boldsymbol{0}
		\end{equation}
		We see in (\ref{Trueexpl}) that it gives an unbiased estimation of $\boldsymbol{Y}$ and $\boldsymbol{\beta^*}$
		but in terms of $\boldsymbol{\beta}$ this estimator is biased:
		\begin{equation}
			\operatorname{E}[\boldsymbol{\hat\beta_1^*}|\boldsymbol{X_1}]=\boldsymbol{\beta_1}+\sum_{j \in I_2}\beta_{j}\boldsymbol{\alpha_j} \textrm{ and }\operatorname{E}[\boldsymbol{\hat\beta_2^*}|\boldsymbol{X_1}]=\boldsymbol{0}
		\end{equation}
		with variance:
		\begin{equation}
			\operatorname{Var}[\boldsymbol{\hat\beta_1^*}|\boldsymbol{X_1}]= (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 )(\boldsymbol{X_1}' \boldsymbol{X_1})^{-1}  \textrm{ and }\operatorname{Var}[\boldsymbol{\hat\beta_2^*}|\boldsymbol{X_1}]= \boldsymbol{0} 
		\end{equation}
		We see that the variance is reduced compared to OLS described in equation (\ref{eqOLS})(no correlations and smaller matrix give better conditioning ) for small values of $\sigma_j$ $i.e.$ strong correlations. So we play on the bias-variance tradeoff, reducing the variance by adding a bias.		
			%There is no theoretical guarantee that our model is better. It's just a compromise between numerical issues caused by correlations for estimation and selection versus increased variability due to structural hypothesis. Therefore we made some simulations to compare both methods (see the end of this paper).
			 \\
	\subsection{Illustration of the tradeoff conveyed by the pretreatment}		  
	 The Mean Squared Error (\textsc{MSE}) on $\boldsymbol{\hat{\beta}}$ is:
	\begin{eqnarray}
		\textsc{MSE}(\boldsymbol{\hat{\beta}}|\boldsymbol{\beta},\boldsymbol{X})&=&\parallel \operatorname{Bias}\parallel_2^2+\operatorname{Tr}(\operatorname{Var}(\boldsymbol{\hat{\beta}})) \\
			&=& 0 + \sigma_Y^2 \operatorname{Tr}((\boldsymbol{X}'\boldsymbol{X})^{-1}) \textrm{ for OLS, and then for the marginal model:} \\
			&=& \parallel\sum_{j \in I_2}\beta_{j}\boldsymbol{\alpha_j} \parallel_2^2 +\parallel \boldsymbol{\beta_2}\parallel^2_2 + (\sigma^2_Y+\sum_{j \in I_2}\sigma^2_{j}\beta_{j}^2 ) \operatorname{Tr}((\boldsymbol{X_1}' \boldsymbol{X_1})^{-1})
	\end{eqnarray}	 
	To better illustrate the bias-variance tradeoff, we look at a simple example with $p=3$ variables. $\boldsymbol{X_1}$ is composed by two independent scaled Gaussian $\mathcal{N}(0,1)$, $\boldsymbol{x_3}=\boldsymbol{x_1}+\boldsymbol{x_2}+\boldsymbol{\varepsilon_3}$ where $\boldsymbol{\varepsilon_3}\sim{\mathcal{N}(\boldsymbol{0},\sigma_3^2\boldsymbol{I_n})}$. We also have $\boldsymbol{\beta}=(1,1,1)$ and $\sigma_Y=10$. Then we observe the theoretical Mean Squared Error (MSE) of the estimator of both OLS and \textsc{CorReg}'s marginal  model for several values of $\sigma_3$ (strength of the sub-regression) and $n$. Figures \ref{MQE1} to \ref{MQE3} show the theoretical MSE evolution with the strength of the sub-regression:
	\begin{equation}
		1-\mathcal{R}^2=\frac{\operatorname{Var}(\boldsymbol{\varepsilon_3)}}{\operatorname{Var}(\boldsymbol{x_3})}=\frac{\sigma_3^2}{\sigma_3^2+2}
	\end{equation}

\begin{figure}[h!]
	\begin{minipage}[l]{.32\linewidth}
			\includegraphics[width=170px]{figures/MQEn15sigmaY10.png} 
			\caption{For $n=15$. Dotted: \textsc{Correg}, plain: OLS}\label{MQE1}
	\end{minipage} \hfill
	\begin{minipage}[c]{.32\linewidth}
			\includegraphics[ width=170px]{figures/MQEn100sigmaY10.png} 
			\caption{For $n=100$. Dotted: \textsc{Correg}, plain: OLS}
	\end{minipage} \hfill
   \begin{minipage}[r]{.32\linewidth}
			\includegraphics[width=170px]{figures/MQEn1000sigmaY10.png} 
			\caption{For $n=1000$. Dotted: \textsc{Correg}, plain: OLS.} \label{MQE3}
   \end{minipage} 
\end{figure} 
It is clear in Figures \ref{MQE1} to \ref{MQE3} that the marginal model is more robust than \textsc{OLS} on $X$. And when sub-regression get weaker ($1-\mathcal{R}^2$ tends to 1) it does not explode until extreme values (sub-regression nearly fully explained by the noise). We also see that the error implied by strong correlations shrinks with the rise of $n$. 
We see that $\sigma_y$ multplies $\operatorname{Tr}(\operatorname{Var}(\boldsymbol{\hat{\beta}}))=\operatorname{Tr}(\operatorname{Var}(\boldsymbol{\hat{\beta}_{1}}))+\operatorname{Tr}(\operatorname{Var}(\boldsymbol{\hat{\beta}_{2}}))$ for both models but for the marginal model $\operatorname{Tr}(\operatorname{Var}(\boldsymbol{\hat{\beta}_{2}}))=0$.
 Thus, as $\operatorname{Tr}(\operatorname{Var}(\boldsymbol{\hat{\beta}_{2}}))=0$ when $\sigma_Y^2$ rises it increases the advantage of \textsc{CorReg} versus \textsc{OLS}. It illustrates the importance of dimension reduction when the model has a strong noise (very usual case on real datasets where true model is not even exactly linear).
